Exploring dimensions of biodiversity in Japanese ferns

### 2020-08-19 ms/SI.Rmd
Update SI
- remove bi plots for raw vs SES biodiv metrics

```{r endemism-restricted, fig.width = 7, fig.height = 8}
# PE (endemic)
# FIXME: cells with zero species should be NA, not zero PE
a <- ggplot(ses_div_ferns_endemic_spatial, aes(fill = pe_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(fill = "PE") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme() +
  theme(
    axis.text.x = element_blank()
  )

# SES PE (endemic)
# FIXME: cells with zero species should be NA, not zero PE
b <- ggplot(ses_div_ferns_endemic_spatial, aes(fill = pe_obs_p_upper)) +
  geom_sf(color = "transparent") +
  labs(fill = "PE p-score") +
  scale_fill_scico(palette = "lajolla") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

# RPE
c <- ggplot(ses_div_ferns_endemic_spatial, aes(fill = rpe_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(fill = "RPE") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_light() +
  theme(
    axis.text.x = element_blank()
  )

# SES RPE
d <- ggplot(ses_div_ferns_endemic_spatial, aes(fill = rpe_obs_z)) +
  geom_sf(color = "transparent") +
  labs(fill = "SES RPE") +
  scale_fill_scico(
    palette = "roma",
    na.value="grey80",
    direction = -1,
    limits = c(
      -get_limit(ses_div_ferns_endemic_spatial, rpe_obs_z, "abs"),
      get_limit(ses_div_ferns_endemic_spatial, rpe_obs_z, "abs")
    )) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme() +
  theme(axis.text.y = element_blank())

# CANAPE, restricted dataset
# set color scheme
canape_cols <- c(
  "neo" = "#E41A1C", # Color brewer set1 red
  "paleo" = "#377EB8", # Color brewer blue
  "mixed" = "#984EA3", # Color brewer purple
  "none" = "#FFFFCC" # Scico lajolla faintest color
)

e <- ses_div_ferns_endemic_spatial %>%
  mutate(endem_type = replace_na(endem_type, "none")) %>%
  ggplot(aes(fill = endem_type)) +
  geom_sf(color = "transparent") +
  labs(fill = "Endemism\ntype") +
  scale_fill_manual(values = canape_cols, breaks = c("neo", "paleo", "mixed", "none"), labels = c("Neo", "Paleo", "Mixed", "Not endemic")) + 
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_light()

a + b + c + d + e +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = 'A')

ggsave(result_file(s_figure("endemism-restricted"), "pdf"), width = 7, height = 8, units = "in")
```

**`r s_figure("endemism-restricted")`**. `r s_figure_cap("endemism-restricted")` **A**, phylogenetic endemism (PE). **B**, Standard effect size (SES) of PE. **C**, relative phylogenetic endemism (RPE). **D**, SES of RPE. **E**, Categorical analysis of neo- and paleo-endemism (CANAPE).

@@

```{r biplot-prep}
# Extract centroids
centroids <- sf::st_centroid(ses_div_ferns_spatial$geometry) %>%
  sf::st_coordinates()

# Define helper function for making a scatterplot
scatterplot <- function(data, x_var, y_var, x_lab, y_lab, label, annotate, loc) {
  p <- ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(alpha = 0.5) +
    labs(
      x = x_lab,
      y = y_lab
    )

  # Add annotation and trendline if annotate = TRUE
  if(annotate & loc == "UL") p <- p +
      annotate("label", x = -Inf, y = Inf, label = label, hjust = 0, vjust = 1, label.size = 0, size = 3, fill = "transparent") +
      geom_smooth(method = "lm", se = FALSE, color = "blue", size = 0.5)

  if(annotate & loc == "UR") p <- p +
      annotate("label", x = Inf, y = Inf, label = label, hjust = 1, vjust = 1, label.size = 0, size = 3, fill = "transparent") +
      geom_smooth(method = "lm", se = FALSE, color = "blue", size = 0.5)

  if(annotate & loc == "LL") p <- p +
      annotate("label", x = -Inf, y = -Inf, label = label, hjust = 0, vjust = 0, label.size = 0, size = 3, fill = "transparent") +
      geom_smooth(method = "lm", se = FALSE, color = "blue", size = 0.5)

  if(annotate & loc == "LR") p <- p +
      annotate("label", x = Inf, y = -Inf, label = label, hjust = 1, vjust = 0, label.size = 0, size = 3, fill = "transparent") +
      geom_smooth(method = "lm", se = FALSE, color = "blue", size = 0.5)

  p
}
```

```{r raw-biplot, fig.width = 8, fig.height = 10.5}
## Biplots of raw metrics ##

# Set up x and y variables as factors so they can be ordered appropriately in the plot
x_vars <-
  c("pd_obs", "rpd_obs", "fd_obs", "rfd_obs", "pe_obs", "rpe_obs", "richness") %>%
  factor() %>%
  fct_relevel(as.character(.))

y_vars <- c("pd_obs", "rpd_obs", "fd_obs", "rfd_obs", "pe_obs", "rpe_obs") %>%
  factor() %>%
  fct_relevel(as.character(.))

# Make dataframe of arguments to feed into plotting function
# - includes all non-duplicated combinations of raw diversity metrics
raw_comparisons <-
  cross_df(
    list(
      x_var = x_vars,
      y_var = y_vars
    ),
    .filter = `==` # filter out exact duplicates
  ) %>%
  # Remove reversed duplicates
  # https://stackoverflow.com/questions/22756392/deleting-reversed-duplicates-with-r
  group_by(grp = paste(pmax(as.character(x_var), as.character(y_var)), pmin(as.character(x_var), as.character(y_var)), sep = "_")) %>%
  arrange(x_var, y_var) %>%
  slice(1) %>%
  ungroup() %>%
  select(-grp) %>%
  arrange(x_var, y_var) %>%
  # Convert x_var and y_var back to character for plotting function
  mutate(across(c(x_var, y_var), as.character)) %>%
  # Add labels
  mutate(
    x_lab = case_when(
      x_var == "pd_obs" ~ "PD",
      x_var == "rpd_obs" ~ "RPD",
      x_var == "fd_obs" ~ "FD",
      x_var == "rfd_obs" ~ "RFD",
      x_var == "pe_obs" ~ "PE",
      x_var == "rpe_obs" ~ "RPE",
      x_var == "richness" ~ "Richness"
    ),
    y_lab = case_when(
      y_var == "pd_obs" ~ "PD",
      y_var == "rpd_obs" ~ "RPD",
      y_var == "fd_obs" ~ "FD",
      y_var == "rfd_obs" ~ "RFD",
      y_var == "pe_obs" ~ "PE",
      y_var == "rpe_obs" ~ "RPE"
    )
  ) %>%
  # Run modified t-test
  mutate(
    test_res = map2(
      .x = x_var, 
      .y = y_var, 
      ~SpatialPack::modified.ttest(ses_div_ferns_spatial[[.x]], ses_div_ferns_spatial[[.y]], centroids)),
    # Extract relevant stats from results
    cor = map_dbl(test_res, pluck("corr")),
    pval = map_dbl(test_res, pluck("p.value"))
    ) %>%
  # Generate label
  mutate(
    cor_lab = glue("r = {round(cor, 3)}"),
    pval_lab = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01 ~ "**",
      pval < 0.05 ~ "*",
      TRUE ~ ""
    ),
    label = glue::glue("{cor_lab}{pval_lab}")
  ) %>%
  # Specify which plots to annotate
  mutate(
    annotate = case_when(
      pval < 0.05 ~ TRUE,
      TRUE ~ FALSE
    ),
    # annotate in upper-right corner for mpd_morph_obs + rpe_obs,
    # upper-left otherwise
    loc = case_when(
      y_var == "mpd_morph_obs" & x_var == "rpe_obs" ~ "UR",
      TRUE ~ "UL"
    )
  ) %>%
  select(x_var, y_var, x_lab, y_lab, label, annotate, loc)

# Make plot
pmap(raw_comparisons, scatterplot, data = ses_div_ferns_spatial) %>%
wrap_plots() +
  plot_layout(ncol = 3)

ggsave(result_file(s_figure("raw-biplot"), "pdf"), width = 8, height = 10.5, units = "in")
```

**`r s_figure("raw-biplot")`**. `r s_figure_cap("raw-biplot")` MPD, mean phylogenetic distance; MMD, mean morphological distance; PD, phylogenetic diversity; PE, phylogenetic endemism; RPE, relative phylogenetic endemism. Correlations between variables analyzed with Dutilleul's modified *t*-test that accounts for spatial structure in the data; *p* value and correlation (r) only shown for *p* < 0.05.

```{r ses-biplot, fig.width = 8, fig.height = 10.5}
## Biplots of SES metrics ##

# Set up x and y variables as factors so they can be ordered appropriately in the plot
x_vars <-
  c("pd_obs_z", "rpd_obs_z", "fd_obs_z", "rfd_obs_z", "pe_obs_z", "rpe_obs_z", "richness") %>%
  factor() %>%
  fct_relevel(as.character(.))

y_vars <- c("pd_obs_z", "rpd_obs_z", "fd_obs_z", "rfd_obs_z", "pe_obs_z", "rpe_obs_z") %>%
  factor() %>%
  fct_relevel(as.character(.))

# Make dataframe of arguments to feed into plotting function
# - includes all non-duplicated combinations of raw diversity metrics
ses_comparisons <-
  cross_df(
    list(
      x_var = x_vars,
      y_var = y_vars
    ),
    .filter = `==` # filter out exact duplicates
  ) %>%
  # Remove reversed duplicates
  # https://stackoverflow.com/questions/22756392/deleting-reversed-duplicates-with-r
  group_by(grp = paste(pmax(as.character(x_var), as.character(y_var)), pmin(as.character(x_var), as.character(y_var)), sep = "_")) %>%
  arrange(x_var, y_var) %>%
  slice(1) %>%
  ungroup() %>%
  select(-grp) %>%
  arrange(x_var, y_var) %>%
  # Convert x_var and y_var back to character for plotting function
  mutate(across(c(x_var, y_var), as.character)) %>%
  # Add labels
  mutate(
    x_lab = case_when(
      x_var == "pd_obs_z" ~ "SES PD",
      x_var == "rpd_obs_z" ~ "SES RPD",
      x_var == "fd_obs_z" ~ "SES FD",
      x_var == "rfd_obs_z" ~ "SES RFD",
      x_var == "pe_obs_z" ~ "SES PE",
      x_var == "rpe_obs_z" ~ "SES RPE",
      x_var == "richness" ~ "Richness"
    ),
    y_lab = case_when(
      y_var == "pd_obs_z" ~ "SES PD",
      y_var == "rpd_obs_z" ~ "SES RPD",
      y_var == "fd_obs_z" ~ "SES FD",
      y_var == "rfd_obs_z" ~ "SES RFD",
      y_var == "pe_obs_z" ~ "SES PE",
      y_var == "rpe_obs_z" ~ "SES RPE"
    )
  ) %>%
  # Run modified t-test
  mutate(
    test_res = map2(
      .x = x_var, 
      .y = y_var, 
      ~SpatialPack::modified.ttest(ses_div_ferns_spatial[[.x]], ses_div_ferns_spatial[[.y]], centroids)),
    # Extract relevant stats from results
    cor = map_dbl(test_res, pluck("corr")),
    pval = map_dbl(test_res, pluck("p.value"))
    ) %>%
  # Generate label
  mutate(
    cor_lab = glue("r = {round(cor, 3)}"),
    pval_lab = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01 ~ "**",
      pval < 0.05 ~ "*",
      TRUE ~ ""
    ),
    label = glue::glue("{cor_lab}{pval_lab}")
  ) %>%
  # Specify which plots to annotate
  mutate(
    annotate = case_when(
      pval < 0.05 ~ TRUE,
      TRUE ~ FALSE
    ),
    # annotate in upper-right corner for mpd_morph_obs + rpe_obs,
    # upper-left otherwise
    loc = case_when(
      y_var == "rpe_obs_z" & x_var == "pe_obs_z" ~ "LR",
      y_var == "fd_obs_z" & x_var == "richness" ~ "UR",
      y_var == "pd_obs_z" & x_var == "richness" ~ "UR",
      TRUE ~ "UL"
    )
  ) %>%
  select(x_var, y_var, x_lab, y_lab, label, annotate, loc)

# Make plot
pmap(ses_comparisons, scatterplot, data = ses_div_ferns_spatial) %>%
wrap_plots() +
  plot_layout(ncol = 3)

ggsave(result_file(s_figure("ses-biplot"), "pdf"), width = 8, height = 10.5, units = "in")
```

**`r s_figure("ses-biplot")`**. `r s_figure_cap("ses-biplot")` MPD, mean phylogenetic distance; MMD, mean morphological distance; PD, phylogenetic diversity; PE, phylogenetic endemism; RPE, relative phylogenetic endemism. Correlations between variables analyzed with Dutilleul's modified *t*-test that accounts for spatial structure in the data; *p* value and correlation (r) only shown for *p* < 0.05.

>>>>

```{r raw-biplot, fig.width = 7, fig.height = 7}
a <- ggplot(biodiv_ferns_spatial, aes(x = richness, y = pd_obs)) +
  geom_point(alpha = 0.5) +
  labs(x = "Richness", y = "PD")

b <- ggplot(biodiv_ferns_spatial, aes(x = richness, y = rpd_obs)) +
  geom_point(alpha = 0.5) +
  labs(x = "Richness", y = "RPD")

c <- ggplot(biodiv_ferns_spatial, aes(x = richness, y = fd_obs)) +
  geom_point(alpha = 0.5) +
  labs(x = "Richness", y = "FD")

d <- ggplot(biodiv_ferns_spatial, aes(x = richness, y = rfd_obs)) +
  geom_point(alpha = 0.5) +
  labs(x = "Richness", y = "RFD")

a + b + c + d +
  plot_layout(ncol = 2, nrow = 2) +
  plot_annotation(tag_levels = 'A')

ggsave(result_file(s_figure("raw-biplot"), "pdf"), width = 7, height = 3, units = "in")
```

**`r s_figure("raw-biplot")`**. `r s_figure_cap("raw-biplot")` **A**, Phylogenetic diversity (PD). **B**, Relative phylogenetic diversity (RPD). **C**, Functional diversity (FD). **D**, Relative functional diversity (RFD).

@@

```{r map-plot-prep}
# Download high-res map of Japan,
# crop to spatial div results area
japan <- rnaturalearth::ne_countries(country = "japan", scale = "large", returnclass = "sf") %>%
  sf::st_crop(sf::st_bbox(biodiv_ferns_endemic_spatial))

# Set CRS
japan_crs <- sf::st_crs(japan)

biodiv_ferns_endemic_spatial <- sf::st_set_crs(biodiv_ferns_endemic_spatial, japan_crs)

# Set color palettes
# Randomization test significance
signif_cols <-
  c(
    "> 0.99" = "#e31a1c", 
    "> 0.975" = "#fb9a99",
    "< 0.01" = "#1f78b4", 
    "< 0.025" = "#a6cee3",
    "not significant" = "#ffffcc"
  )

# CANAPE
canape_cols <-
  c(
    "neo" = "#e31a1c", 
    "paleo" = "#1f78b4",
    "mixed" = "#6a51a3",
    "super" = "#4a1486",
    "not significant" = "#ffffcc"
  )
```

```{r endemism-restricted, fig.width = 5, fig.height = 5}
ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_endemic_spatial, aes(fill = endem_type), color = "transparent", alpha = 0.8) +
  scale_fill_manual(values = canape_cols) +
  map_theme() +
  theme(legend.title = element_blank())

ggsave(result_file(s_figure("endemism-restricted"), "pdf"), width = 5, height = 5, units = "in")
```

**`r s_figure("endemism-restricted")`**. `r s_figure_cap("endemism-restricted")` 'paleo' indicates paleoendemic areas (those with an overabundance of long branches); 'neo' indicates neodendemic areas (those with an overabundance of short branches); 'mixed' indicates significantly endemic sites that have neither an overabundance of short nor long branches; 'super' indicates highly significantly endemic sites that have neither an overabundance of short nor long branches. For details, see Methods.
